---
title: "Police Shootings"  
author: "Ryan S Dunn and Derek Lankeauz"
date: "11/28/2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

### Import the relevent data sets for EDA and model development

```{r}
#import the data sets for EDA
library(readxl)

#import the police shootings since 2015 data 
police_post2015 <- read.csv("~/Documents/USD MS-ADS/Applied Data Mining 502/Final Project/PoliceShootings_post2015.txt")

#import supplementary income, poverty, race, and high school graduation data for data blending/joining 
median_income <- read_excel("~/Documents/USD MS-ADS/Applied Data Mining 502/Final Project/MedianHouseholdIncome2015.xlsx")
povery_level <- read_excel("~/Documents/USD MS-ADS/Applied Data Mining 502/Final Project/PercentagePeopleBelowPovertyLevel.xlsm")
race_city <- read_excel("~/Documents/USD MS-ADS/Applied Data Mining 502/Final Project/ShareRaceByCity.xlsm")
hs_grad <- read_excel("~/Documents/USD MS-ADS/Applied Data Mining 502/Final Project/PercentOver25CompletedHighSchool.xlsm")
```

#### Import the necessary libraries
```{r}
library(ggplot2)
library(dplyr)
```

#### Develop a median income data frame to join onto the police_shootings dataframe

```{r}
#develop a median income object to join onto the police shootings data frame
income_df <- data.frame((median_income))

#change data types as needed
income_df$Median.Income <- as.numeric(income_df$Median.Income)
income_df$Geographic.Area <- as.factor(income_df$Geographic.Area)

#aggregate the median income via the median median income of each state
income_table <- aggregate(x = income_df$Median.Income,
          by = list(income_df$Geographic.Area),
          FUN =  median)

#save the income_table as a data frame and convert the names of the columns
income_table <- as.data.frame(income_table)
income_table <- rename(income_table, "State" = "Group.1")
income_table <- rename(income_table, "Median.Income" = "x")

#view the developed object
income_table
```

#### Develop a povery rate data frame to join onto the police_shootings dataframe

```{r}
#develop a poverty rate object to join onto the police shootings data frame
pr_df <- data.frame(povery_level)

#change the data types as needed
pr_df$poverty_rate  <- as.numeric(pr_df$poverty_rate)
pr_df$Geographic.Area <- as.factor(pr_df$Geographic.Area)

#aggregate the poverty rate via the median poverty rate of each state
pr_table <- aggregate(x = pr_df$poverty_rate,
          by = list(pr_df$Geographic.Area),
          FUN =  median)

#save the pr_table as a data frame and convert the names of the columns
pr_table <- as.data.frame(pr_table)
pr_table  <- rename(pr_table, "State" = "Group.1")
pr_table <- rename(pr_table, "Median.Below.Poverty" = 'x')

#view the object
pr_table 
```

#### Develop a percent of population over 25 years old that has graduated from high school data frame to join onto the police_shootings dataframe

```{r}
#develop a hs rate object to join onto the police shootings data frame
hs_df <- data.frame(hs_grad)

#change the data types as needed
hs_df$Geographic.Area <- as.factor(hs_df$Geographic.Area)
hs_df$percent_completed_hs <- as.numeric(hs_df$percent_completed_hs)

#aggregate the hs_df as a data frame and covert the names of the columns
hs_table <- aggregate( x = hs_df$percent_completed_hs, 
                       by = list(hs_df$Geographic.Area),
                       FUN = median)

hs_table <- as.data.frame(hs_table)
hs_table <- rename(hs_table, "State" = "Group.1")
hs_table <- rename(hs_table, "Over.25.Grad.Rate" = "x")

#view the object
hs_table

```

#### Join the developed data frame data onto the police shootings data 

```{r}
#develop the final_df object from the police shootings and left joined data from the developed objects
final_df <- left_join(police_post2015, pr_table, by = c("state" = "State"))
final_df <- left_join(final_df, income_table, by = c("state" = "State"))
final_df <- left_join(final_df, hs_table, by = c("state" = "State"))

#create the regional column data frame 
head(final_df)
```
#### Add in a region area by state 
```{r}
final_df <- final_df %>% mutate(Region =
                     case_when(state == 'AL' ~ 'Southeast',
                               state == 'AK' ~ 'West',
                               state == 'AZ' ~ 'Southwest',
                               state == 'AR' ~ 'Southeast',
                               state == 'CA' ~ 'West',
                               state == 'CO' ~ 'West',
                               state == 'CT' ~ 'Northeast',
                               state == 'DE' ~ 'Northeast',
                               state == 'DC' ~ 'Southeast',
                               state == 'FL' ~ 'Southeast',
                               state == 'GA' ~ 'Southeast',
                               state == 'GU' ~ 'West',
                               state == 'HI' ~ 'West',
                               state == 'ID' ~ 'West',
                               state == 'IL' ~ 'Midwest',
                               state == 'IN' ~ 'Midwest',
                               state == 'IA' ~ 'Midwest',
                               state == 'KS' ~ 'Midwest',
                               state == 'KY' ~ 'Southeast',
                               state == 'LA' ~ 'Southeast',
                               state == 'ME' ~ 'Northeast',
                               state == 'MD' ~ 'Northeast',
                               state == 'MA' ~ 'Northeast',
                               state == 'MI' ~ 'Midwest',
                               state == 'MN' ~ 'Midwest',
                               state == 'MS' ~ 'Southeast',
                               state == 'MO' ~ 'Midwest',
                               state == 'MT' ~ 'West',
                               state == 'NE' ~ 'Midwest',
                               state == 'NV' ~ 'West',
                               state == 'NH' ~ 'Northeast',
                               state == 'NJ' ~ 'Northeast',
                               state == 'NM' ~ 'Southwest',
                               state == 'NY' ~ 'Northeast',
                               state == 'NC' ~ 'Southeast',
                               state == 'ND' ~ 'Midwest',
                               state == 'OH' ~ 'Midwest',
                               state == 'OK' ~ 'Southwest',
                               state == 'OR' ~ 'West',
                               state == 'PA' ~ 'Northeast',
                               state == 'PR' ~ 'Southeast',
                               state == 'RI' ~ 'Northeast',
                               state == 'SC' ~ 'Southeast',
                               state == 'SD' ~ 'Midwest',
                               state == 'TN' ~ 'Southeast',
                               state == 'TX' ~ 'Southwest',
                               state == 'UT' ~ 'West',
                               state == 'VA' ~ 'Southeast',
                               state == 'VT' ~ 'Northeast',
                               state == 'WA' ~ 'West',
                               state == 'WV' ~ 'Southeast',
                               state == 'WI' ~ 'Midwest',
                               state == 'WY' ~ 'West')) 
```

```{r}
final_df <- final_df %>% mutate(Armed.Flag =
                     case_when(armed == 'undertermed' ~ '0',
                               armed == 'unarmed' ~ '0',
                               armed == 'NA' ~ '0'))
#repalce all NA's in the Armed.Flag with a 1 flag
final_df[is.na(final_df)] <- 1 
```

```{r}
#develop an attribute that depicts if a person is a minority or not
final_df <- final_df %>% mutate(Is.Minority = 
                                  case_when(race == 'W' ~ '0'))
final_df[is.na(final_df)] <- '1'
```

```{r}
#display a contigency table to review that the output of the above mutation is correct
a <- table(final_df$race, final_df$Is.Minority)
a
```
## Fleeing Contigency Tables 
```{r}
#table for armed
armed_table <- table(final_df$Armed.Flag, final_df$race)
armed_table
```

```{r}
round(prop.table(armed_table, margin = 2)*100,1)
```
## Begin Exploratory Data Analysis

```{r}
summary(final_df)
```

#### Deaths by Gender bar chart
```{r}
gender_summary_post <- table(police_post2015$gender)

gender_post <- barplot(gender_summary_post[order(gender_summary_post, decreasing = FALSE)],
                main = "Deaths by Gender Post-2015",
                xlab = 'Gender',
                ylab = 'Deaths')
text(gender_post,  + 2000 , gender_summary_post, font=1)
```

```{r}
#police shootings by gender box chart 
ggplot(final_df, aes(gender)) + geom_bar(aes(fill= Is.Minority)) + ggtitle("Stacked Bar Chart by Gender with Minority Overlay") +
  scale_fill_manual(values = c("blue","lightblue")) + labs(x="Gender", y="Count of Deaths")
```
### Deaths by Region 
```{r}
cont_table_region <- table(final_df$Region)
cont_table_region
```

```{r}
#bar chart with deaths by region with minority overlay
ggplot(final_df, aes(Region)) + geom_bar(aes(fill= Is.Minority)) + ggtitle("Stacked Bar Chart by Region with Minority Overlay") +
  scale_fill_manual(values = c("blue","lightblue")) + labs(x="Region", y="Count of Deaths")

region_summary_table <- table(final_df$Region)
```

```{r}
cont_table_region <- table(final_df$Region)
cont_table_region
```
#### Deaths by Age histogram 
```{r}
#histogram of age post-2015
ggplot(data = police_post2015, aes(age)) + geom_histogram(binwidth = 5, color='black'  ,fill ='lightblue') + ggtitle("Histogram of Police Shooting Victims by Age") +labs(x ='Age of Shooting Victim', y = 'Count of Victims')
```

#### Deaths by Race bar chart
```{r}
#create a table of the deaths by race
race_summary_post <- table(police_post2015$race)

#develop the bar chart in decreasing order
race_post <- barplot(race_summary_post[order(race_summary_post, decreasing = TRUE)],
               main = "Deaths by Race Post-2015",
               xlab = 'Race',
               ylab = 'Deaths')

# W = White, B = Black, H = Hispanic, A = Asian, Empty = Unknown, N = Native American, O = Other 
```

```{r}
#return the vector of only the deaths by race category
race_summary_post
```

#### Develop contingency tables of police shootings by race, and associated proportions/percentages of whole

```{r}
#contigency tables and percentages of shooting by race
cont_table_race <- table(police_post2015$race)
prop_table_race <- prop.table(cont_table_race) 
perc_table_race <- prop.table(cont_table_race) * 100

race_table <- rbind(cont_table_race, prop_table_race, perc_table_race)
rownames(race_table) <- c("Count","Proportion","Percentage")
race_table
#race_table["Count", "B"]
```

#### From the total U.S. Population statistics in 2019, develop the race proporitions of the U.S. and determine the associated distributions of police shootings by race relative to race proportion in the U.S. 

```{r}
#from: https://www.visualcapitalist.com/visualizing-u-s-population-by-race/ -- retrieve U.S. Population and Demographic data

#estimated U.S. Populations as of 2019
total_pop <- 328239523

#estimated U.S. race demographic proportions
white_pop <- .601 * total_pop
black_pop <- .122 *total_pop
hisp_pop <- .185 *total_pop
asian_pop <- .056 * total_pop
other_pop <- 100 - white_pop -black_pop - hisp_pop - asian_pop

#develop an object by race of the count of deaths by the population proportion
white_prop <- (race_table["Count","W"] / white_pop) * 100
black_prop <- (race_table["Count","B"] / black_pop) * 100
hisp_prop <-  (race_table["Count","H"] / hisp_pop)  * 100
asian_prop <- (race_table["Count","A"] / asian_pop) * 100 
```

```{r}
#print the developed race proportions of deaths by police shooting
print(black_prop)
print(hisp_prop)
print(white_prop)
print(asian_prop)
```

### Scatterplot of HS Grad Rate, Median Below Povery by Region

```{r}
ggplot(data=final_df) + 
  geom_point(mapping = aes( x = Median.Below.Poverty, y = Over.25.Grad.Rate, color = Region)) + ggtitle("Scatterplot") + xlab("Median Below Poverty Rate") + ylab("% of Population over 25 -HS Grads") 
```
```{r}
head(final_df)
```
### Scatterplot of Lat $ Long by Region

```{r}
ggplot(data=final_df) + 
  geom_point(mapping = aes( x = longitude, y = latitude, color = Region))
```

### Race and Region Contigency Tables

```{r}
race_region_cont <- table(final_df$race,final_df$Region)
race_region_cont
```

```{r}
round(prop.table(race_region_cont, margin = 2)*100,1)
```

### Bar Chart of Total Police Shootings with Race Overlay

```{r}
ggplot(final_df, aes(race)) + geom_bar(aes(fill=Region)) + ggtitle("Stacked Bar Chart of Race and Region") +  scale_fill_manual(values=c("darkgrey","grey","lightblue","blue","darkblue"))
```

```{r}
#normalized bar chart
ggplot(final_df, aes(race)) + geom_bar(aes(fill=Region) , position = "fill") + ggtitle("Normalized Stacked Bar Chart of Race and Region") +  scale_fill_manual(values=c("darkgrey","grey","lightblue","blue","darkblue"))
```
### Histograms of Age with Race Overlay

```{r}
#histogram of age with race underlay
ggplot(final_df, aes(age)) + geom_histogram(aes(fill=race), color="black", binwidth = 5) + ggtitle("Histogram of Age with Race Overlay") +labs( x = 'Age of Shooting Victim' , y='Count of Victims') +  scale_fill_manual(values=c("darkgrey","grey","lightblue","blue","darkblue","cyan","cyan4"))
```

```{r}
#histogram of age with race underlay
ggplot(final_df, aes(age)) + geom_histogram(aes(fill=race), color="black", binwidth = 5, position = "fill") + ggtitle("Normalized Histogram of Age with Race Overlay") +labs( x = 'Age of Shooting Victim' , y='Count of Victims') +  scale_fill_manual(values=c("darkgrey","grey","lightblue","blue","darkblue","cyan","cyan4"))
```

```{r}
ggplot(final_df, aes(Region)) + geom_bar(aes(fill=race)) + ggtitle("Stacked Bar Chart of Race and Region") + scale_fill_manual(values=c("darkgrey","grey","lightblue","blue","darkblue","cyan","cyan4"))
```

```{r}
ggplot(data =final_df, mapping = aes(x=race, y = age)) + geom_boxplot() + ggtitle("Boxplot of Age of Police Victims by Race")
```



